---
title: "Testing Taxonomic Features"
output:
  github_document:
    df_print: tibble
editor_options: 
  chunk_output_type: console
---

```{r global_options, include=FALSE}
source(here::here("common.R"))
```

Here is a workflow for testing taxonomic features using the Segata/Huttenhower tool, LEfSe.
```{r}
library(tidyverse)
library(yingtools2)
library(phyloseq)
library(ggtree)
```


First create a phyloseq object with the predictors of interest, by adding variables from `cid.patients`.
I am planning to test the microbiome changes associated with the antibiotic metronidazole.
```{r}
phy <- cid.phy
s <- get.samp(phy) %>% left_join(cid.patients,by="Patient_ID") %>%
  mutate(metronidazole=ifelse(metronidazole,"metronidazole","no metronidazole"))
sample_data(phy) <- s %>% set.samp()
```

Use `lda.effect` to test the effect of the variable `metronidazole`. First it will perform Kruskal-Wallis testing of abundances for each taxonomic feature.
Specify `subclass="sex"` to perform additional Wilcoxon testing of various subsets across `sex`. Finally, a linear discriminant analysis (LDA) is performed on taxonomic features selected by the above testing steps. From this, an effect size is calculated. 
```{r}
lda <- lda.effect(phy,class="metronidazole",subclass="sex")
```


A common way to display the data is to plot as horizontal bars:
```{r fig.height=4, fig.width=10}
lda.plot(lda)
```

You can also view the data in a cladogram, using `lda.clado`. 
```{r fig.height=9, fig.width=10}
lda.clado(lda)
```







