---
title: "Testing Taxonomic Features"
output:
  github_document:
    df_print: tibble
editor_options: 
  chunk_output_type: console
---

```{r global_options, include=FALSE}
source(here::here("common.R"))
```

Here is a workflow for testing taxonomic features using the Segata/Huttenhower tool, LEfSe.
```{r}
library(tidyverse)
library(yingtools2)
library(phyloseq)
library(ggtree)
```


First create a phyloseq object with the predictors of interest, by adding variables from `cid.patients`.
I am planning to test the microbiome changes associated with the antibiotic metronidazole.
```{r}
phy <- cid.phy
s <- get.samp(phy) %>% left_join(cid.patients,by="Patient_ID") %>%
  mutate(metronidazole=ifelse(metronidazole,"metronidazole","no metronidazole"))
sample_data(phy) <- s %>% set.samp()
```

Use `lda.effect` to test the effect of the variable `metronidazole`. First it will perform Kruskal-Wallis testing of abundances for each taxonomic feature.
Specify `subclass="sex"` to perform additional Wilcoxon testing of various subsets across `sex`. Finally, a linear discriminant analysis (LDA) is performed on taxonomic features selected by the above testing steps. From this, an effect size is calculated. 
```{r}
lda <- lda.effect(phy,class="metronidazole",subclass="sex")
lda
```


A common way to display the data is to plot as horizontal bars:
```{r fig.height=4, fig.width=10}
lda.plot <- lda %>% filter(pass) %>%
  mutate(lda=if_else(as.numeric(factor(direction))==2,-lda,lda)) %>%
  arrange(lda) %>%
  mutate(taxonomy=fct_inorder(taxonomy))

ggplot(lda.plot,aes(x=taxonomy,y=lda,fill=direction)) + geom_col() + coord_flip()
```

You can also view the data in a cladogram. 
```{r fig.height=9, fig.width=9}
# Add columns to the LDA table listing separate taxonomic levels (Kingdom, Phylum, ...) in full form. 
# This will lead to correct labelling of all nodes.
lvls <- rank_names(phy)
for (i in 1:length(lvls)) {
  lda[[lvls[i]]] <- str_split(lda.tbl$taxonomy,"\\|") %>% map_chr(~str_c(.[1:i],collapse="|"))
}
# Use the newly create columns to create a new phylo object representing the taxonomic hierarchy.
# Then plot with ggtree.
lefse.phy <- as.phylo.formula2(as.formula(paste("~",paste(lvls,collapse="/"))),data=lda)
gt <- ggtree(lefse.phy,layout="circular")
# This recursive function calculates the range of y values for all descendants of a given node.
get.children.yrange <- function(node,gd) {
  hits <- gd$node[gd$parent==node]
  if (length(hits)==0 | node %in% hits) {
    return(gd$y[gd$node==node])
  } else {
    return(unlist(lapply(hits,get.children.yrange,gd)))
  }
}
gd <- gt$data %>% left_join(lda,by=c("label"="taxonomy")) %>%
  mutate(y.range=lapply(node,get.children.yrange,cur_data()),
  ymin=map_dbl(y.range,min),
  ymax=map_dbl(y.range,max),xmin=x,xmax=9+7-x,
  ymid=(ymin+ymax)/2,
  xtext=xmax-0.5,
  angle.label=scales::rescale(ymid,from=range(y),to=c(0,360)),
  angle.label=if_else(is.between(angle.label,0,180),angle.label-90,angle.label+90),
  short.label=map_chr(str_split(label,"\\|"),last))

gt + geom_point(data=gd,aes(size=log.max),color="dark gray",fill="gray",shape=21) +
  geom_point2(data=filter(gd,pass),aes(fill=direction,size=log.max),shape=21) +
  geom_rect(data=filter(gd,pass),aes(xmin=xmin,xmax=xmax,ymin=ymin,ymax=ymax,fill=direction),color="dark gray",alpha=0.2) +
  geom_text(data=filter(gd,pass),aes(x=xtext,y=ymid,label=short.label,angle=angle.label)) + 
  theme(legend.position="right")

```







