---
title: "Phylogenetic Trees"
output:
  github_document:
    df_print: tibble
editor_options: 
  chunk_output_type: console
---

```{r global_options, include=FALSE}
source(here::here("common.R"))
```

Here I am describing my workflow to create plots of phylogenetic trees. To make phylogenetic trees in R, the `ggtree` package is reasonably good and flexible for our purposes.
```{r}
library(tidyverse)
library(yingtools2)
library(phyloseq)
library(ggtree)
```


Tree data is most commonly stored as a `phylo` object in R. These can be stored within `phyloseq` objects as a part of the data. 
Using `cid.phy` data as an example, the `phylo` object can be extracted using `phy_tree()` function.  

This works with the `phylo` object, which is the most common way to store phylogenetic trees in R.

```{r}
tr <- phy_tree(cid.phy)
tr
```

Use the `ggtree()` function to create a `ggtree` object. This is essentially a custom `ggplot` object which can be further modified and tweaked with additional steps.
```{r}
ggtree(tr)
```

You can tweak the appearance by specifying `layout`:
```{r,fig.show="hold", out.width="33%"}
ggtree(tr,layout="circular") + ggtitle("circular")
ggtree(tr,layout="dendrogram") + ggtitle("dendrogram")
ggtree(tr,layout="ape") + ggtitle("ape")
```

You can assign useful taxonomic data to the `ggtree` object by using the `%<+%` operator, which is specific to the `ggtree` package. 
Like other `ggplot` objects, there is actually a data frame stored within the `ggtree` object which can be used for plotting. 
I like to save this as a separate object (`gd`), which makes downstream customizations a little easier. If you examine this, you'll start to understand what `ggtree` is doing.
```{r}
gt <- ggtree(tr) %<+% get.tax(cid.phy)
gd <- gt$data
gd
```

If I wanted to plot points at each tip, I can add the layer in usual ggplot2 fashion. 
```{r}
gt + geom_point(aes(color=Phylum))
```

...but because we only want the tips to be labelled (not the nodes), use `geom_point2`, which simply allows subsetting of the data in a convenient way. 
```{r}
gt + geom_point2(aes(subset=isTip,color=Phylum))
```

Let's cover a few more details. 
I'd like to plot where tips are at genus level (the one above was at OTU/ASV level), so use `phy.collapse` to create a genus-level tree. 
There are too many Phylum level colors, collapse the small ones into "Other Phyla".
```{r}
phy.genus <- cid.phy %>% phy.collapse(taxranks=c("Kingdom","Phylum","Class","Order","Family","Genus"))
tr <- phy_tree(phy.genus)
gt <- ggtree(tr) %<+% get.tax(phy.genus)
gd <- gt$data %>% mutate(Phylum=fct_lump_n(Phylum,4))
gt + geom_point2(data=gd,aes(subset=isTip,color=Phylum))
```

Use `hilight.clade()` to add layers representing clades, using the ggtree data. There are a lot parameters you can tweak to customize exact how you want it to look.
```{r}
gt + geom_point2(data=gd,aes(subset=isTip,color=Phylum)) +
  hilight.clade(gt,Phylum,"Actinobacteria",ymin=101,ymax=155,fill.color="purple",alpha=0.1) +
  hilight.clade(gd,Phylum,"Bacteroidetes",ymin=300,ymax=328,fill.color="light blue",alpha=0.3)
```



