---
title: "Phylogenetic Trees"
output:
  github_document:
    df_print: tibble
editor_options: 
  chunk_output_type: console
---

```{r global_options, include=FALSE}
source(here::here("common.R"))
```

Here I am describing my workflow to create plots of phylogenetic trees. To make phylogenetic trees in R, the `ggtree` package is reasonably good and flexible for our purposes.
```{r}
library(tidyverse)
library(yingtools2)
library(phyloseq)
library(ggtree)
```


Tree data is most commonly stored as a `phylo` object in R. These can be stored within `phyloseq` objects as a part of the data. 
Using `cid.phy` data as an example, the `phylo` object can be extracted using `phy_tree()` function.  

This works with the `phylo` object, which is the most common way to store phylogenetic trees in R.

```{r}
tr <- phy_tree(cid.phy)
tr
```

Use the `ggtree()` function to create a `ggtree` object. This is essentially a custom `ggplot` object which can be further modified and tweaked with additional steps.
```{r}
ggtree(tr)
```

You can tweak the appearance by specifying `layout`:
```{r,fig.show="hold", out.width="33%"}
ggtree(tr,layout="circular") + ggtitle("circular")
ggtree(tr,layout="dendrogram") + ggtitle("dendrogram")
ggtree(tr,layout="ape") + ggtitle("ape")
```

You can assign useful taxonomic data to the `ggtree` object by using the `%<+%` operator, which is specific to the `ggtree` package. 
Like other `ggplot` objects, there is actually a data frame stored within the `ggtree` object which can be used for plotting. 
I like to save this as a separate object (`gd`), which makes downstream customizations a little easier. If you examine this, you'll start to understand what `ggtree` is doing.
```{r}
gt <- ggtree(tr) %<+% get.tax(cid.phy)
gd <- gt$data
gd
```

If I wanted to plot points at each tip, I can add the layer in usual ggplot2 fashion. 
```{r}
gt + geom_point(aes(color=Phylum))
```

...but because we only want the tips to be labelled (not the nodes), use `geom_point2`, which simply allows subsetting of the data in a convenient way. 
```{r}
gt + geom_point2(aes(subset=isTip,color=Phylum))
```

Let's cover a few more details. 
I'd like to plot where tips are at genus level (the one above was at OTU/ASV level), so use `phy.collapse` to create a genus-level tree. 
There are too many Phylum level colors, collapse the small ones into "Other Phyla".
```{r}
phy.genus <- cid.phy %>% phy.collapse(taxranks=c("Kingdom","Phylum","Class","Order","Family","Genus"))
tr <- phy_tree(phy.genus)
gt <- ggtree(tr) %<+% get.tax(phy.genus)
gd <- gt$data %>% mutate(Phylum=fct_lump_n(Phylum,4))
gt + geom_point2(data=gd,aes(subset=isTip,color=Phylum))
```

Use `hilight.clade()` to add layers representing clades, using the ggtree data. There are a lot parameters you can tweak to customize exact how you want it to look.
```{r}
gt + geom_point2(data=gd,aes(subset=isTip,color=Phylum)) +
  hilight.clade(gt,Phylum,"Actinobacteria",ymin=101,ymax=155,fill.color="purple",alpha=0.1) +
  hilight.clade(gd,Phylum,"Bacteroidetes",ymin=300,ymax=328,fill.color="light blue",alpha=0.3)
```

```{r eval=FALSE, include=FALSE}

geom_hilight <- function(mapping = NULL, data = NULL, stat = "identity",
                         position = "identity", show.legend = NA,
                         inherit.aes = TRUE, 
                         na.rm = TRUE, opt1 = "option1", opt2 = ~option2, 
                         ...) {
  layer(
    geom = GeomHilight,
    stat = stat,
    data = data,
    mapping = mapping,
    position = position,
    params = list(na.rm = na.rm, opt1 = opt1, opt2 = opt2),
    inherit.aes = inherit.aes,
    check.aes = TRUE,
    check.param = TRUE,
    show.legend = show.legend,
    key_glyph = NULL
  )
}


GeomHilight <- ggproto("GeomHilight", Geom,
                    draw_panel = function(self, data, panel_params, coord, opt1, opt2) {
                      points <- data %>% select(x, y, PANEL, group, shape, colour, size, fill, alpha, stroke)
                      cols <- data %>% #GeomCol$setup_data(panel_params) %>%
                        select(PANEL, ymin, ymax, xmin, xmax, colour, fill, size, linetype, alpha)
                      lines <- data %>% select(x, y, PANEL, group, colour, size, linetype, alpha)
                      text <- data %>% select(x, y, label, PANEL, group, colour=fontcolour, size=fontsize,
                                              angle, hjust, vjust,
                                              alpha, family, fontface, lineheight)
                      segments <- data %>% group_by(PANEL,colour,group,size,linetype,alpha) %>%
                        summarize(xend=max(x),x=min(x)) %>%
                        mutate(y=-5,yend=-5) %>%
                        ungroup()
                      grid::gList(
                        GeomPoint$draw_panel(points, panel_params, coord),
                        GeomLine$draw_panel(lines, panel_params, coord),
                        GeomRect$draw_panel(cols, panel_params, coord),
                        GeomCol$draw_panel(cols, panel_params, coord),
                        GeomText$draw_panel(text, panel_params, coord),
                        GeomSegment$draw_panel(segments, panel_params, coord),
                        NULL
                      )
                    },
                    draw_group = function(self, data, panel_params, coord, opt1, opt2) {},
                    draw_key = draw_key_point,
                    required_aes = c("x", "y", "label"),
                    default_aes = aes(colour = "black", shape = 19, size = 1, fill = NA, alpha = NA, stroke = 0.5,
                                      linetype=1, 
                                      angle = 0, hjust = 0.5, vjust = 0.5, family = "", fontface = 1, 
                                      fontsize = 4, fontcolour="red",
                                      lineheight = 1.2),

                    non_missing_aes = NULL,
                    optional_aes = NULL,
                    extra_params = c("na.rm", "opt1", "opt2"),
                    setup_data = GeomCol$setup_data #creates xmin/xmax
                    # draw_layer = xxxx,
)




```


```{r eval=FALSE, include=FALSE}
library(ggplot2)
library(tidyverse)

geom_xxxx <- function(mapping = NULL, data = NULL, stat = "identity",
                         position = "identity", show.legend = NA,
                         inherit.aes = TRUE, 
                         na.rm = TRUE, opt1 = "option1", opt2 = ~option2, 
                         ...) {
  layer(
    geom = GeomXXXX,
    stat = stat,
    data = data,
    mapping = mapping,
    position = position,
    params = list(na.rm = na.rm, opt1 = opt1, opt2 = opt2),
    inherit.aes = inherit.aes,
    check.aes = TRUE,
    check.param = TRUE,
    show.legend = show.legend,
    key_glyph = NULL
  )
}


GeomXXXX <- ggproto("GeomXXXX", Geom,
                    draw_panel = function(self, data, panel_params, coord, opt1, opt2) {
                      points <- data %>% select(x, y, PANEL, group, shape, colour, size, fill, alpha, stroke)
                      cols <- data %>% #GeomCol$setup_data(panel_params) %>%
                        select(PANEL, ymin, ymax, xmin, xmax, colour, fill, size, linetype, alpha)
                      lines <- data %>% select(x, y, PANEL, group, colour, size, linetype, alpha)
                      text <- data %>% select(x, y, label, PANEL, group, colour=fontcolour, size=fontsize,
                                              angle, hjust, vjust,
                                              alpha, family, fontface, lineheight)
                      segments <- data %>% group_by(PANEL,colour,group,size,linetype,alpha) %>%
                        summarize(xend=max(x),x=min(x)) %>%
                        mutate(y=-5,yend=-5) %>%
                        ungroup()
                      grid::gList(
                        GeomPoint$draw_panel(points, panel_params, coord),
                        GeomLine$draw_panel(lines, panel_params, coord),
                        GeomRect$draw_panel(cols, panel_params, coord),
                        GeomCol$draw_panel(cols, panel_params, coord),
                        GeomText$draw_panel(text, panel_params, coord),
                        GeomSegment$draw_panel(segments, panel_params, coord),
                        NULL
                      )
                    },
                    draw_group = function(self, data, panel_params, coord, opt1, opt2) {},
                    draw_key = draw_key_point,
                    required_aes = c("x", "y", "label"),
                    default_aes = aes(colour = "black", shape = 19, size = 1, fill = NA, alpha = NA, stroke = 0.5,
                                      linetype=1, 
                                      angle = 0, hjust = 0.5, vjust = 0.5, family = "", fontface = 1, 
                                      fontsize = 4, fontcolour="red",
                                      lineheight = 1.2),

                    non_missing_aes = NULL,
                    optional_aes = NULL,
                    extra_params = c("na.rm", "opt1", "opt2"),
                    setup_data = GeomCol$setup_data #creates xmin/xmax
                    # draw_layer = xxxx,
)


ggplot(mtcars,aes(x=mpg,y=disp,label=cyl),fill="gray",size=1) + geom_xxxx()

ggplot(mtcars,aes(x=mpg,y=disp,label=cyl)) + geom_text()

GeomPoint %>% {c(.$required_aes,names(.$default_aes),"group","PANEL")}
GeomBar %>% {c(.$required_aes,names(.$default_aes),"group","PANEL")} 
GeomText %>% {c(.$required_aes,names(.$default_aes),"group","PANEL")}
GeomRect %>% {c(.$required_aes,names(.$default_aes),"group","PANEL")}
GeomSegment %>% {c(.$required_aes,names(.$default_aes),"group","PANEL")}
GeomSmooth %>% {c(.$required_aes,names(.$default_aes),"group","PANEL")}
GeomBar %>% {c(.$required_aes,names(.$default_aes),"group","PANEL")}
GeomCol %>% {c(.$required_aes,names(.$default_aes),"group","PANEL")}
GeomViolin %>% {c(.$required_aes,names(.$default_aes),"group","PANEL")}
GeomBoxplot %>% {c(.$required_aes,names(.$default_aes),"group","PANEL")}
GeomDotplot %>% {c(.$required_aes,names(.$default_aes),"group","PANEL")}
GeomBlank %>% {c(.$required_aes,names(.$default_aes),"group","PANEL")}
GeomCustomAnn %>% {c(.$required_aes,names(.$default_aes),"group","PANEL")}
GeomCrossbar %>% {c(.$required_aes,names(.$default_aes),"group","PANEL")}
GeomErrorbar %>% {c(.$required_aes,names(.$default_aes),"group","PANEL")}


# library(ggdebug)
# GeomCol$setup_params
# ggdebug(ggplot)
# ggdebug(GeomRect$draw_panel)
# ggdebug(GeomText$draw_panel)
# ggdebug(GeomDotplot$draw_panel)
# ggdebug(GeomXXXX$draw_group)
# ggundebug_all()



```






